---
title: Closeness Index - One-run games++
author: ~
date: '2018-06-17'
slug: closeness-index-one-run-games
categories: []
tags: []
output:
  blogdown::html_page
---

## Introduction

The [last post](../../../../2018/06/15/record-in-one-run-games/) looked at records in one-run games. It turns out that [Baseball Musings](http://www.baseballmusings.com/?p=48169) offered a alternative way of thinking about it some 8 years ago. The overarching idea here is that one-run games are kind of a proxy for tight games - but there's a difference between a game that's not close, and then the bullpen blows a lead and it ends up at 1-run, versus a game that's close the whole way. The second type is qualitatively different and tells us more about how strategy might factor in.

To reemphasize what the technical idea is, you compute a closeness index for a game by taking the average of the absolute value of the score differential at the end of each half inning. So there's a lower limit for a regulation 9-inning game of 1/18 = 0.056, i.e. when it's tied the entire game up until the bottom of the ninth and then the home team wins it by one.

This post will use this definition for closeness index and do some exploratory analysis.

## libraries

``` {r, message=FALSE}
library(RPostgres)
library(DBI)
library(dplyr)
library(ggplot2)
library(ggrepel)
```

## data

I'll use the retrosheet event-level data. This is overkill in a way since I only need scores at the end of half-innings, but it's the best way I know how to get that.

In the post where Baseball Musings introduced closeness, he looked exclusively at 1974+, where the records are 100% complete. I'm going to look at 1955+ since I'm looking mainly for outliers and overall trends and am not overly concerned with completeness - I generally believe the maxim of don't let the perfect be the enemy of the good.

Connect to the database

``` {r}
conn <- dbConnect(RPostgres::Postgres(), 
                  password=Sys.getenv("PSQL_PASS"), 
                  user=Sys.getenv("PSQL_USER"), 
                  port=Sys.getenv("PSQL_PORT"), 
                  dbname='retrosheet')
```


Grab data since 1955, filtering to only include end of innings

``` {r}
  ev = dbGetQuery(conn, "select * from event where year_id > 1954 and inn_end_fl = \'t\'")
```

### closeness per inning

This function processes the data to:

* figure out the score at each half-inning

* make a duplicate of each record so that there is a row for the home team as "the team" and the away team as "the opponent", and visa versa

* add a column for "closeness_idx", the score differential at the half-inning

``` {r}
process_data = function(ev) {
  ev2 = ev %>% 
    mutate(home_score = home_score_ct + ifelse(bat_home_id==1, event_runs_ct, 0), 
           away_score = away_score_ct + ifelse(bat_home_id==0, event_runs_ct, 0))
  evA = ev2 %>% 
    select(game_id, inn_ct, year_id, bat_home_id, 
           team_id=home_team_id, opp_id=away_team_id, 
           team_score=home_score, opp_score=away_score)
  evH = ev2 %>% 
    select(game_id, inn_ct, year_id, bat_home_id,
           team_id=away_team_id, opp_id=home_team_id, 
           team_score=away_score, opp_score=home_score)
  
  evX = dplyr::bind_rows(evA, evH) %>% 
    mutate(close_idx = abs(team_score - opp_score))

}
```

Process the event data

``` {r}
evX = process_data(ev)
```


### closeness per game

This function groups the half-inning data by game, adding a column to indicate which team won.


``` {r}

get_wins = function(evX) {
  win_df = evX %>% 
    group_by(game_id) %>% 
    mutate(mx = max(inn_ct)) %>% 
    filter(inn_ct==mx) %>% 
    arrange(-bat_home_id) %>% 
    mutate(ridx= row_number()) %>% 
    filter(ridx <= 2) %>% mutate(is_win= team_score > opp_score) %>% 
    select(game_id, team_id, is_win)
  
  win_df %>% inner_join(evX, by=c("game_id", "team_id")) %>% 
    group_by(game_id, year_id, team_id) %>% 
    mutate(closeness_score=mean(close_idx)) %>% ungroup()
}

```


``` {r}
ev_win_df = get_wins(evX)
ev_win_df %>% head(3) %>% as.data.frame()
```

Aggregate by game, computing the closeness index as the mean absolute value of the score differential 

``` {r}
game_win_df = ev_win_df %>% 
  group_by(game_id, team_id, is_win, year_id) %>% 
  summarise(closeness_idx=mean(close_idx)) %>% 
  ungroup()
game_win_df %>% head(3) %>% as.data.frame()
```

## exploratory analysis

The distribution of closeness index

``` {r, fig.height=8, fig.width=8, fig.units="in"}
p = game_win_df %>% ggplot(aes(x=closeness_idx)) + 
  geom_histogram(binwidth = 0.5) + 
  theme_minimal(base_size = 16)
print(p)
```


### the "most close" games

This collapses the per-game closeness (one record for each team) into one record per game, and orders them by closeness index. The `game_id` can be copy pasted to look up the game on baseball reference, e.g., [BAL195705192](https://www.baseball-reference.com/boxes/BAL/BAL195705192.shtml)

``` {r}
game_win_df %>% group_by(game_id) %>% 
  summarise(closeness_idx=mean(closeness_idx)) %>%
  arrange(closeness_idx) %>% 
  filter(closeness_idx < 0.032) %>% 
  knitr::kable()
```

### the "least close" games

Same for the highest closeness indexes, i.e. the "least close" games

``` {r}
game_win_df %>% group_by(game_id) %>% 
  summarise(closeness_idx=mean(closeness_idx)) %>%
  arrange(-closeness_idx) %>% 
  filter(closeness_idx > 12.6) %>% 
  knitr::kable()
```

## seasonal trends

Anecdotally based on the tables above, the close games tend to come from low-scoring seasons and the least close ones from high-scoring seasons. This highlights the fairly self-evident fact that score differential, and hence closeness, depends on the run environment. This shows the closeness index by season, averaged over all games in a season.


``` {r, fig.height=8, fig.width=8, fig.units="in"}
p = game_win_df %>% group_by(year_id) %>% 
  summarise(closeness_idx = mean(closeness_idx)) %>% 
  ggplot(aes(x=year_id, y=closeness_idx)) + 
  geom_point(size=0.5) + 
  geom_line() + theme_minimal(base_size = 16)
print(p)
```

## Team Records

### fracvtio of one-run games

The ultimate goal of this analysis was to look at records in close games, as an analogy to the records in one-run games analyses. To do that, we need to account for the run environment, as shown above. To get an idea of how many games we should consider close and where to set the cutoff, this shows the fraction of all games that are one-run games.

``` {r, fig.height=8, fig.width=8, fig.units="in"}
p = ev_win_df %>% 
  group_by(game_id, year_id, team_id) %>% 
  summarise(team_score=max(team_score), 
            opp_score = max(opp_score)) %>%
  ungroup() %>% 
  mutate(is1 = abs(team_score - opp_score) == 1) %>% 
  group_by(year_id) %>% 
  summarise(frac1 = mean(is1)) %>% 
  ggplot(aes(x=year_id, y=frac1)) + 
  geom_line() + theme_minimal(base_size = 16)
print(p)
```

We see that historically "one-run games" occur something like 30% of the time. Here I take the lowest 30% of closeness index as definition of "close games" and see which teams had the most wins and highest win percentage

First define a function to change the definition and output tables of most wins and highest win percentage

``` {r}
close_wins_tables = function(game_wins_df, close_definition, nteam=15) {

  close_percentile_df = game_wins_df %>% 
    mutate(dummy=1) %>% 
    group_by(year_id) %>% 
    arrange(year_id, closeness_idx) %>% 
    mutate(running = cumsum(dummy), n=n(), c_frac=running / n) %>%
    ungroup() %>% 
    mutate(is_close = c_frac <= close_definition) %>% 
    group_by(year_id, team_id) %>% 
    summarise(w=sum(is_win), g=n(), 
              w_close = sum(is_close * is_win), 
              g_close = sum(is_close), 
              wpct_close = w_close/g_close) 
  
  list(w_table = 
         close_percentile_df %>% 
         arrange(-w_close) %>% 
         head(nteam),
       wpct_table = 
         close_percentile_df %>% 
         arrange(-wpct_close) %>% 
         head(nteam)
       )

}
```

With the general function defined, try using the 30% closest games for each season

``` {r}
close_definition = 0.3
tables = close_wins_tables(game_win_df, close_definition)
# ordered by wins
knitr::kable(tables$w_table)
# ordered by win percentage
knitr::kable(tables$wpct_table)
```

Let's tighten that up to 20th percentile

``` {r}
close_definition = 0.2
tables = close_wins_tables(game_win_df, close_definition)
knitr::kable(tables$w_table)
knitr::kable(tables$wpct_table)
```

And 10th percentile,

``` {r}
close_definition = 0.1
tables = close_wins_tables(game_win_df, close_definition)
# ordered by wins
knitr::kable(tables$w_table)
# ordered by win percentage
knitr::kable(tables$wpct_table)
```


## non-close games

We can try the same, looking at games that aren't close. This doesn't really reflect any strategy but is more just a fun piece of trivia to see which teams had good records in blowouts. 

The function definition is almost identical to the one for close games, except we pick games greater than some threshold versus games less than some threshold.

``` {r}
blowout_wins_tables = function(game_wins_df, close_definition, nteam=15) {

  close_percentile_df = game_wins_df %>% 
    mutate(dummy=1) %>% 
    group_by(year_id) %>% 
    arrange(year_id, closeness_idx) %>% 
    mutate(running = cumsum(dummy), n=n(), c_frac=running / n) %>%
    ungroup() %>% 
    mutate(is_close = c_frac >= close_definition) %>% 
    group_by(year_id, team_id) %>% 
    summarise(w=sum(is_win), g=n(), 
              w_close = sum(is_close * is_win), 
              g_close = sum(is_close), 
              wpct_close = w_close/g_close) %>% 
    rename(w_blowout=w_close, wpct_blowout=wpct_close)
  
  list(w_table = 
         close_percentile_df %>% 
         arrange(-w_blowout) %>% 
         head(nteam),
       wpct_table = 
         close_percentile_df %>% 
         arrange(-wpct_blowout) %>% 
         head(nteam)
       )

}
```

``` {r}
blowout_definition = 0.8
tables = blowout_wins_tables(game_win_df, blowout_definition)
# ordered by wins
knitr::kable(tables$w_table)
# ordered by win percentage
knitr::kable(tables$wpct_table)
```

Ratchet it up to 90th percentile

``` {r}
blowout_definition = 0.9
tables = blowout_wins_tables(game_win_df, blowout_definition)
# ordered by wins
knitr::kable(tables$w_table)
# ordered by win percentage
knitr::kable(tables$wpct_table)
```


## Records in close vs non-close games

The main point output of the look at [one-run games](../../../../2018/06/15/record-in-one-run-games/) was a graph of win percentage in one-run vs non-one-run games. This graph reproduces that, using top 20th percentile of closeness as the replacement for one-run games.


``` {r}
close_definition = 0.2

plot_df = close_percentile_df = game_win_df %>% 
  mutate(dummy=1) %>% 
  group_by(year_id) %>% 
  arrange(year_id, closeness_idx) %>% 
  mutate(running = cumsum(dummy), n=n(), c_frac=running / n) %>%
  ungroup() %>% 
  mutate(is_close = c_frac <= close_definition) %>% 
  group_by(year_id, team_id) %>% 
  summarise(w=sum(is_win), g=n(), 
            w_close = sum(is_close * is_win), 
            g_close = sum(is_close), 
            wpct_close = w_close/g_close, 
            wpct_other = (w-w_close)/(g - g_close)) %>% 
  ungroup()
```


``` {r, fig.height=8, fig.width=8, fig.units="in"}
# the base plot
p = plot_df %>% 
  ggplot(aes(x=wpct_other, y=wpct_close)) + 
  geom_point() + 
  theme_minimal(base_size = 16) + 
  geom_hline(yintercept = 0.5, size=0.5, color='steelblue') + 
  geom_vline(xintercept = 0.5, size=0.5, color='steelblue') + 
  labs(x="Win Pct. - Non close games", y="Win Pct. - close games")
``` 

Label the ones that are on the edges

``` {r}
records_quads = plot_df %>% 
  mutate(r2=(wpct_other - 0.5)**2 + (wpct_close - 0.5)**2, 
         quad=as.integer(wpct_other >= 0.5) + 2 * as.integer(wpct_close >= 0.5))

lab_df = records_quads %>% 
  group_by(quad) %>% 
  arrange(-r2) %>% 
  mutate(r2_rank=row_number()) %>% 
  filter(r2_rank <= 10) %>% 
  ungroup() %>% 
  mutate(name=paste(team_id, year_id))
```


``` {r, fig.height=8, fig.width=8, fig.units="in"}
# base plot with outliers labelled
p2 = p + 
  geom_text_repel(data = lab_df, aes(x=wpct_other, y=wpct_close-wpct_other, label=name))
print(p2)

``` 


## 

``` {r, fig.height=8, fig.width=8, fig.units="in"}
# the base plot
p = plot_df %>% 
  ggplot(aes(x=wpct_other, y=wpct_close-wpct_other)) + 
  geom_point() + 
  theme_minimal(base_size = 16) + 
  geom_hline(yintercept = 0.0, size=0.5, color='steelblue') + 
  geom_vline(xintercept = 0.5, size=0.5, color='steelblue') + 
  labs(x="Win Pct. - Non close games", y="Win Pct.Diff. - (close - non-close)")
``` 

Label the ones that are on the edges

``` {r}
records_quads = plot_df %>% 
  mutate(r2=(abs(wpct_other - 0.5))**1.2 + (abs(wpct_close-wpct_other))**1.4, 
         quad=as.integer(wpct_other >= 0.5) + 2 * as.integer(wpct_close - wpct_other >= 0.0))

lab_df = records_quads %>% 
  group_by(quad) %>% 
  arrange(-r2) %>% 
  mutate(r2_rank=row_number()) %>% 
  mutate(i1 = r2_rank <= 12, 
         i2 = wpct_other>=0.65 & abs(wpct_close-wpct_other)<0.1, 
         i3=wpct_close-wpct_other>0.29,
         iX = i1+i2+i3) %>% 
  filter(iX > 0) %>%
  ungroup() %>% 
  mutate(name=paste(team_id, year_id))

```


``` {r, fig.height=8, fig.width=8, fig.units="in"}
# base plot with outliers labelled
tdf = plot_df %>% mutate(wp = w/g*162 - 81, wins_above_81=as.integer(round(wp/10)*10), wpx=ifelse(abs(wp)>=9, wp, 0))

p2 = p + geom_point(data=tdf, aes(color=wins_above_81)) + scale_color_gradient2(high  = 'orange1', low = 'steelblue') + geom_text_repel(data = lab_df, aes(x=wpct_other, y=wpct_close-wpct_other, label=name))
print(p2)

``` 